% =====================================================================
% Modeling Risk -- Problem Set 2 (2026)
% Full report linking theory, code (PS2/) and numerical results
% Compile with:  pdflatex -shell-escape main.tex   (twice for refs)
% =====================================================================
\documentclass[11pt,a4paper]{article}

% ---------- Encoding & language ----------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\usepackage{lmodern}
\usepackage{microtype}

% ---------- Math ----------
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}
\usepackage{dsfont}

% ---------- Layout ----------
\usepackage[a4paper,margin=2.4cm]{geometry}
\usepackage{setspace}
\usepackage{parskip}
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{siunitx}
\sisetup{
  detect-all,
  group-separator={\,},
  output-decimal-marker={.}
}

% ---------- Floats ----------
\usepackage{graphicx}
\graphicspath{{../figures/}{./figures/}}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}

% ---------- Refs & links ----------
\usepackage[colorlinks=true,
            linkcolor=blue!55!black,
            citecolor=blue!55!black,
            urlcolor=blue!55!black]{hyperref}
\usepackage{cleveref}

% ---------- Code listings ----------
\usepackage{xcolor}
\usepackage{listings}
\definecolor{codebg}{RGB}{248,248,248}
\definecolor{codekey}{RGB}{0,0,160}
\definecolor{codestr}{RGB}{160,40,40}
\definecolor{codecom}{RGB}{0,110,0}
\lstdefinestyle{py}{
  language=Python,
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{codekey}\bfseries,
  stringstyle=\color{codestr},
  commentstyle=\color{codecom}\itshape,
  showstringspaces=false,
  backgroundcolor=\color{codebg},
  frame=single,
  rulecolor=\color{black!25},
  framesep=4pt,
  framerule=0.4pt,
  breaklines=true,
  numbers=left,
  numberstyle=\tiny\color{black!50},
  numbersep=6pt,
  xleftmargin=2.2em,
}
\lstset{style=py}

% ---------- Theorem-like ----------
\newtheorem{prop}{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{defn}{Definition}

% ---------- Macros ----------
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\Cor}{\operatorname{Corr}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\indic}{\mathds{1}}
\newcommand{\given}{\,|\,}
\newcommand{\Pphi}{\Phi}
\newcommand{\pphi}{\phi}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\code}[1]{\texttt{\small #1}}
\newcommand{\file}[1]{\texttt{\small\textcolor{blue!55!black}{#1}}}

% ---------- Title ----------
\title{\bfseries Modeling Risk --- Homework 2026\\[2pt]}
\author{Pedro José Piqueras Martínez\\\small Master in Economics and Data Science}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
\noindent
This report develops the theoretical answer to the homework of
\emph{Modeling Risk}. Every mathematical derivation is cross-referenced to
the Python implementation under \file{PS2/}, so that each printed number and
figure can be regenerated by running the corresponding \code{main.py}.
\end{abstract}

\tableofcontents
\bigskip\hrule

% ---------------------------------------------------------------------
\section{Introduction and notation}\label{sec:intro}
% ---------------------------------------------------------------------
A \emph{copula} is the joint distribution function of a random vector with uniform
marginals. By Sklar's theorem any joint distribution $F$ with marginals $F_1,\ldots,F_n$
can be written
\begin{equation}\label{eq:sklar}
F(x_1,\ldots,x_n)=C\!\bigl(F_1(x_1),\ldots,F_n(x_n)\bigr),
\end{equation}
Sklar's theorem is the central tool of the work: it allows us to model the
dependence structure (the copula $C$) separately from the univariate behaviour
of each return ($F_i$). The conditional CDF of the copula
\begin{equation}\label{eq:cond-cdf}
C(u_1\given u_2)=\frac{\partial}{\partial u_2}C(u_1,u_2)
\end{equation}
is the building block of all the simulation algorithms used below: drawing
$V_1,V_2$ i.i.d. uniform on $(0,1)$ and inverting $V_1=C(U_1\given V_2)$ in
$U_1$ produces a sample with the desired copula. We adopt the notation
$\Phi$ for the standard Normal CDF, $\phi$ for its density, $t_\nu$ for the
Student-$t$ CDF with $\nu$ degrees of freedom, and $\eta_t=\Phi^{-1}(u_t)$
for the Normal transform of a uniform.

\paragraph{Project layout.}
The whole numerical work is implemented in a single Python project
\file{PS2/} organised by exercise:
\begin{center}
\small
\begin{tabular}{ll}
\toprule
Folder & Content\\
\midrule
\file{exercise1/} & Bivariate simulation, portfolio statistics, IFM/CML estimation\\
\file{exercise2/} & GARCH/GJR + DCC on BTC/ETH/ADA returns\\
\file{exercise3/} & NM-copula conditional CDF and $q$-quantile inversion\\
\file{exercise4/} & Black--Scholes pricing + option-based portfolio simulation\\
\file{exercise5/} & Clayton DGP, copula MLE, Kupiec backtest\\
\file{config.json} & All numerical parameters consumed by every \code{main.py}\\
\file{figures/} & Generated figures and saved tables\\
\file{report/} & This LaTeX document\\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Reproducibility.}
The file \code{main.py} runs all excercises and the report. 
However, each exercise can be launched from its own \code{main.py}. Each code loads the global
\file{config.json} and writes its outputs to \file{figures/}.
Random seeds are fixed in the configuration file. The exact tables and figures
shown below were obtained by running, from \file{PS2/},
\begin{lstlisting}[language=bash,numbers=none]
# for all excercises
python -m main
# for individual excercises
python -m exercise1.main
python -m exercise2.main
python -m exercise3.main
python -m exercise4.main
python -m exercise5.main
\end{lstlisting}

% ---------------------------------------------------------------------
\input{sections/exercise1.tex}
\input{sections/exercise2.tex}
\input{sections/exercise3.tex}
\input{sections/exercise4.tex}
\input{sections/exercise5.tex}
% ---------------------------------------------------------------------

\section{Concluding remarks}\label{sec:conclusion}

\begin{enumerate}
\item Exercise~\ref{sec:ex1} shows how a single dependence parameter can produce
      radically different sample dependences (Clayton's lower-tail cluster,
      survival Clayton's upper-tail cluster, the elliptical patterns of Gaussian/$t$,
      and the near-independence of FGM). The IFM estimator recovers the true
      dependence parameter to three decimals on $10^5$ observations; the same
      estimator on the 0.6/0.4 Clayton mixture infers a t-copula with a tiny
      $\hat\nu\approx 2.17$, signaling the strong tail dependence the Gaussian
      copula cannot accommodate.
\item Exercise~\ref{sec:ex2} confirms that crypto returns exhibit the standard
      stylised facts (heavy tails, mild leverage, $\alpha+\beta\to 1$). DCC
      under all three marginal specifications (EDF, GARCH-N, GJR-N) produces
      essentially the same dynamic correlation paths; switching from a
      Gaussian to a Student-$t$ copula raises the in-sample log-likelihood
      substantially and yields $\hat\nu\approx 4.4$, validating the heavy-tail
      multivariate dependence.
\item Exercise~\ref{sec:ex3} provides a numerically inversion of the NM
      conditional copula and visualises the asymmetry generated by mixing
      two Gaussian components with mixed-sign correlations.
\item Exercise~\ref{sec:ex4} reproduces the slide-pages 16--19 results: the
      first-leg statistics (\code{cc}, \code{pp}) are invariant to the copula
      because they involve only one underlying, while the diagonal strategies
      (\code{cc2}, \code{pp2}) react sharply to the joint distribution
      assumption.
\item Exercise~\ref{sec:ex5} reproduces, with a Clayton-only DGP, the central
      message of Berger (2016): \emph{copula misspecification translates
      directly into VaR-coverage failures}. The Clayton model (correctly
      specified) achieves nominal coverage; Gaussian and $t$ copulas reject
      the Kupiec test repeatedly at the 99\% level even at moderate dependence.
\end{enumerate}

\bigskip
\noindent\textbf{References.}
\begin{itemize}\setlength\itemsep{0pt}
\item Berger, T. (2016). \emph{Forecasting value-at-risk using time varying copulas
      and EVT return distributions}. Economic Modelling, 58.
\item Course slides: \emph{Copula} and \emph{Option-based portfolios}.
\item Matlab reference programs \file{ccK\_simuRet2.m}, \file{BS.m},
      \file{simuCC\_Normal.m}, \file{F1.m}, \file{F2.m}.
\end{itemize}

\end{document}
